---
title: altSEE panels - illustrated for proteomics data analysis
author:
- name: Lieven Clement
  affiliation: 
  - Ghent University.
  email: lieven.clement@ugent.be
  
date: "`r BiocStyle::doc_date()`"
package: altSEE
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Panel universe}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{altSEE}
  %\VignetteKeywords{GeneExpression, proteomics, Visualization, QualityControl, GUI}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
library("BiocStyle")
```

```{r, eval=!exists("SCREENSHOT"), include=FALSE}
SCREENSHOT <- function(x, ...) knitr::include_graphics(x, dpi = NA)
```

# Overview

The `r Biocpkg("iSEE")` package provides a general and flexible framework for interactively exploring `SummarizedExperiment` objects.
However, in many cases, more specialized panels are required for effective visualization of specific data types.
The `altSEE` package extends the `iSEE` interactive visualisation framework
to support exploration of data stored in *alternative
experiments* (`altExp`) within a `SingleCellExperiment` (SCE) object.

This vignette illustrates `altSEE` on mass spectrometry-based proteomics. 
Particularly we will use the **protein** to link protein abundances in the main 
experiment that were obtained upon summarisation of their corresponding peptides 
that are stored in the altExp.

We first load in the package:

```{r}
library("iSEEu")
library("altSEE")
library("QFeatures")
library("msqrob2")
library("SingleCellExperiment")
library("scater")
library("BiocFileCache")
```

## Overview 

- Import MS-based proteomics data
- Data processing using `QFeatures`
- Differential abundance analysis using `msqrob2`
- Launch an `altSEE` app to explore the DA proteins and their corresponding 
  peptide intensities, interactively.
  
# Differential abundance

To demonstrate the use of these panels we will perform a differential abundance
analysis on  the Shen data containing 5 spike-in conditions (a, b, c, d and e) 
PRIDE identifierPXD003881 (Shen et al. 2018).
The data is used in the first chapter of our [msqrob2book](https://statomics.github.io/msqrob2book/01-basics.html).
Here we will only use spike-in conditions (a-c) for illustration purposes. 

```{r}
bfc <- BiocFileCache()
peptideFile <- bfcrpath(bfc,"https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/shen/peptides.txt")
```

The quantitative columns in maxQuant start with "Intensity." (note the dot!).

```{r}
peptides <- data.table::fread(peptideFile, check.names = TRUE, integer64 = "double")
(quantCols <- grep("Intensity[.]", names(peptides), value = TRUE))
```

We now can import these data using the readQFeatures function. The peptide level expression data will be imported into R as an instance of class QFeatures named cptac with an assay named peptides. We also use the fnames argument to set the row-names of the peptides assay to the peptide sequences.

```{r}
qf <- readQFeatures(peptides, 
                    quantCols = quantCols, 
                    name = "peptides", 
                    fnames = "Sequence")
```

Below we update the sample (column) annotations to encode the two conditions, A and B, and the original sample numbers.

```{r}
qf$condition <- gsub("Intensity.(.).", "\\1", quantCols)
qf$sampleId <- gsub("Intensity.(..)", "\\1",quantCols)
colData(qf)
```

Here, we will only use first three conditions for illustration purposes.

```{r}
qf <- qf[,qf$condition %in% c("a", "b", "c"),]
colData(qf)
```

## Data processing

We will now perform the data processing and as in `msqrob2` vignette in the 
msqrob2book. 

1. Replace missing values by `NA`.
2. Perform log2-transformation
3. Peptide filtering
4. Normalisation
5. Summarization to protein level

```{r warning=FALSE}
qf <- zeroIsNA(qf, "peptides") # 1. convert 0 to NA

qf <- logTransform(qf, base = 2, i = "peptides", name = "peptides_log") # 2.

qf <- filterFeatures(
qf, ~ Proteins != "" & ## 3. Remove failed protein inference
  !grepl(";", Proteins)) ##  3. Remove protein groups
qf <- filterFeatures(qf, ~ Reverse != "+" & ## 3. Remove decoys
                          Potential.contaminant != "+") ## 3. Remove contaminants
qf <- filterNA(qf, i = "peptides_log", pNA = .67) ## 3. Remove peptides that were picked up in less than 4 samples: pNA = 1-4/12. 

qf <- sweep( #Subtract log2 norm factor column-by-column (MARGIN = 2) 
  qf,
  MARGIN = 2,
  STATS = nfLogMedian(qf,"peptides_log"),
  i = "peptides_log",
  name = "peptides_norm"
) ## 4. Normalisation

qf <- aggregateFeatures(qf,
    i = "peptides_norm", 
    fcol = "Proteins", 
    na.rm = TRUE,
    name = "proteins"
) ## 5. summarisation to protein level
```

## DA at the protein level 

We model the data using msqrob2 and infer differential expression for all 
pairwise comparisons. 

```{r warning=FALSE}
model <- ~ condition
qf <- msqrob(
    qf,  
    i = "proteins",
    formula = model,
    robust = TRUE)

(allHypotheses <- createPairwiseContrasts(
  model, 
  colData(qf), 
  "condition"
))

(L <- makeContrast(
  allHypotheses,
  parameterNames = 
    model.matrix(model, colData(qf)) |>
    colnames())
)

#Rename columns L
colnames(L) <- c("b-a","c-a","c-b")

qf <- hypothesisTest(
  qf,
  i = "proteins",
  contrast = L)
```

Note, that we assessed three different null hypothesis for each protein: 

1. There is no differential abundances between condition b and a
2. There is no differential abundances between condition c and a
3. There is no differential abundances between condition c and b


## DA at the peptide level 

For illustration purposes, i.e. the VolcanoPlot panel for AltExps, 
we also perform the modelling and inference at the peptide level. 

We typically would skip this step in a real analysis, unless we also would like 
to infer DA for post translational modifications. 

We will use the same model and contrasts so we do not need to define these again. 

```{r warning=FALSE}
qf <- msqrob(
    qf,  
    i = "peptides_norm",
    formula = model,
    robust = TRUE)

qf <- hypothesisTest(
  qf,
  i = "peptides_norm",
  contrast = L)
```

## Conversion of QFeatures in to SingleCellExperiment

We now convert the `QFeatures` object to a `SingleCellExperiment` object to 
visualize the data with iSEE and explore both the protein level data as well as
their corresponding peptide intensities that were picked up using the custom 
altSEE panels.

1. We get the protein set with colData, convert it to a 
   SingleCellExperiment object and add a reduced dimensionality plot. 
2. Unpack inference columns 

```{r}
sce <- getWithColData(qf, i ="proteins") |> 
   as("SingleCellExperiment") |> 
   runMDS(exprs_values = 1) # 1. 
rowData(sce) <- rowData(sce) |> 
  as.data.frame() |> 
  dplyr::select(is.atomic) #2. 
```

Add altExps with peptide data. 

1. We get the peptide sets, convert them to a 
   SingleCellExperiment object and add a reduced dimensionality plot. 
2. Unpack inference columns 

```{r}
setNames <- names(qf)
setNames <- setNames[setNames != "proteins"]
for (i in setNames)
{
altExp(sce,i) <- qf[[i]] |> 
     as("SingleCellExperiment") |> 
   runMDS(exprs_values = 1) # 1. 
rowData(altExp(sce,i)) <- rowData(altExp(sce,i)) |> 
  as.data.frame() |> 
  dplyr::select(is.atomic) #2.
}
```

## launch iSEE

```{r}
rdp <- ReducedDimensionPlot(PanelWidth = 6L)
ardp <- AltReducedDimensionPlot(PanelWidth = 6L,
                                Experiment = "peptides_norm")
vp <- VolcanoPlot(PanelWidth = 6L)
avp <- AltVolcanoPlot(PanelWidth = 6L,
                      Experiment = "peptides_norm")
rdt <- RowDataTable(RowSelectionSource = "VolcanoPlot1", PanelWidth = 6L)
ardt <- AltRowDataTable(RowSelectionSource = "AltVolcanoPlot1", 
                        Experiment = "peptides_norm",
                           PanelWidth = 6L)
mlfap <- LinkedFeaturesAssayPlot(
  SelectionExperiment = "(Main)",
  LookupColumn = "(rowname)",
  Experiment = "(Main)",
  AltAssay = "proteins",
  MapColumn = "(rowname)",
  YAxisFeatureSource = "RowDataTable1",
  PlotType = "Scatter + lines",
  XAxis = "Column data",
  XAxisColumnData = "sampleId",
  PanelWidth = 6L
  )

alfap <- LinkedFeaturesAssayPlot(
  YAxisFeatureSource = "RowDataTable1",
  SelectionExperiment = "(Main)",
  LookupColumn = "Proteins",
  Experiment = "peptides_norm",
  PlotType = "Scatter + lines",
  XAxis = "Column data",
  XAxisColumnData = "sampleId",
  MapColumn = "Proteins",
  PanelWidth = 6L
  )

initial_panels <- list(rdp, ardp, vp, avp, rdt, ardt, mlfap, alfap)
```


Launch the app 

```{r launch-app, eval=FALSE}
app <- iSEE(
    sce,
    initial = initial_panels
)

if (interactive()) {
    shiny::runApp(app, launch.browser = TRUE)
}
```


```{r, echo=FALSE}
SCREENSHOT("screenshots/altExpApp.png")
```

Note that you can

1. Select the proteins in the row data table using a part of their name or by 
   selecting features in the volcanoplot
2. Look at the results of another protein by either selecting another protein in 
   the row data table or in the dropdown menus of both linked feature plots
3. adjust the volcano plot for other contrasts e.g. c-a and c-b by selecting the 
   corresponding p-value and log2 fold changes in the volcano plot panel. 
4. also use the Alt row data table to select features and the peptide altExp as 
   the selection source in both panels. You can then select DA peptidoforms in 
   the Alt volcano plot, and visualise the peptidoform and its corresponding 
   protein in their corresponding linked feature assay plots by selecting a 
   peptidofrom from the AltExp row data table.

Note that you can also use the AltFeatureAssayPlot panels if you want to 
visualise features from the AltExp (or the Main experiment) without the use of links, i.e. as the altExp counterpart of FeatureAssayPlots for the main experiment. 