---
title: "Visualising paired scRNA-seq and scATAC-seq data with altSEE"
author:
  - name: "Lieven Clement"
    affiliation: "statOmics, Ghent University"
date: "`r Sys.Date()`"
package: altSEE
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{Visualising paired scRNA-seq and scATAC-seq data with altSEE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment  = "#>"
)
```

# Introduction

The `altSEE` package extends the `iSEE` interactive visualisation framework
to support exploration of data stored in *alternative
experiments* (`altExp`) within a `SingleCellExperiment` (SCE) object.
The key idea is that a "main" modality (e.g. RNA expression) and a
"secondary" modality measured on the same cells (e.g. chromatin accessibility,
surface proteomics) can be linked through a shared feature identifier —
most naturally, a **gene symbol** — so that selecting a gene in the RNA
panel immediately pulls up the related features from the secondary modality.

In the mass spectrometry-based proteomics vignette included with `altSEE`, 
the link is the **protein** that maps proteins to there corresponding peptides.

This vignette demonstrates the same concept for **single-cell multiomics**:
the main experiment contains scRNA-seq gene-level counts, while the `altExp`
holds scATAC-seq chromatin-accessibility peaks annotated to their nearest gene 
or vice versa. 

Selecting a gene of interest in the RNA panel therefore reveals all ATAC peaks
in its vicinity, enabling direct visual comparison of chromatin openness and
transcript abundance across cell types.

We use a subset of the 10X Genomics PBMC Multiome dataset  distributed via the
`SingleCellMultiModal` Bioconductor package.

## Overview

- Load paired scRNA-seq / scATAC-seq data from `SingleCellMultiModal`
  and restructure it as an SCE with an `altExp`.
- Annotate ATAC peaks with the nearest gene symbol using
  `GenomicRanges` and `EnsDb.Hsapiens.v86`.
- Run lightweight QC, normalisation and dimensionality reduction on
  both modalities.
- Launch an `altSEE` app that links gene expression to chromatin
  accessibility interactively.

## Loading packages

```{r libraries}
suppressPackageStartupMessages({
library(altSEE)            # the package we are showcasing
library(iSEE)              # underlying interactive framework

library(SingleCellMultiModal)
library(MultiAssayExperiment)
library(SingleCellExperiment)

library(scran)
library(scater)
library(scuttle)
library(GenomeInfoDb)        # <-- adds keepSeqlevels, seqlevelsStyle, etc.
library(GenomicRanges)
library(EnsDb.Hsapiens.v86)

library(BiocParallel)
})
set.seed(42)
```

# Data


A subset of the data for 500 cells.  

```{r load-mae, cache=TRUE}
data(mae)
mae
```

The object is a `MultiAssayExperiment` containing two `SingleCellExperiment`
objects: `rna` (36 549 genes × 499 cells) and `atac` (108 344 peaks ×
499 cells).
All cells are matched — each barcode in `rna` corresponds to the same barcode
in `atac`.

```{r inspect-mae}
# Dimensions of each modality
lapply(experiments(mae), dim)

# Cell-type labels provided by 10X R&D
table(mae$celltype)
```
```{r}
sce_rna  <- mae[["rna"]]
sce_atac <- mae[["atac"]]
```

# Peak–gene annotation

ATAC peak row names carry genomic coordinates (`chr1:827203-827972`) rather
than gene names.
To enable gene-centric linking in altSEE, we annotate each peak with the
**nearest gene symbol** using `GenomicRanges` and the human Ensembl annotation
database `EnsDb.Hsapiens.v86` (GRCh38).

We also record the *distance to the nearest transcription start site* (TSS)
and whether the peak falls within the gene body.
These annotations are stored in `rowData(sce_atac)` and drive the altSEE
linking logic.

```{r annotate-peaks, cache=TRUE}
edb <- EnsDb.Hsapiens.v86

# 1. Parse peak coordinates into a GRanges object
peak_gr <- GRanges(rownames(sce_atac))

# Ensure chromosome style matches between peaks and EnsDb
# (EnsDb uses "1", "2", …; 10X peaks use "chr1", "chr2", …)
seqlevelsStyle(peak_gr) <- "Ensembl"  # strips "chr" prefix

#2. Retrieve gene-level GRanges from EnsDb 
genes_gr <- genes(edb, columns = c("gene_name", "gene_biotype"))

# Only keep chromosomes present in BOTH genes_gr and peak_gr
std_chroms <- seqlevels(genes_gr)[!grepl("_|\\.", seqlevels(genes_gr))]
std_chroms <- intersect(std_chroms, seqlevels(peak_gr))  # <-- add this line

genes_gr   <- keepSeqlevels(genes_gr, std_chroms, pruning.mode = "coarse")
peak_gr    <- keepSeqlevels(peak_gr,  std_chroms, pruning.mode = "coarse")

# 3. Find nearest gene for each peak 
nearest_idx      <- nearest(peak_gr, genes_gr)
rowData(sce_atac)$nearestGene <- genes_gr$gene_name[nearest_idx]

# 4. Distance from peak midpoint to TSS 
tss_gr <- resize(genes_gr, width = 1, fix = "start")
dist_hits <- distanceToNearest(peak_gr, tss_gr)
rowData(sce_atac)$distToTSS <- mcols(dist_hits)$distance

# 5. Flag promoter peaks (within 2 kb of TSS) 
rowData(sce_atac)$isPromoter <- rowData(sce_atac)$distToTSS <= 2000

# 6. Preview annotation
head(as.data.frame(rowData(sce_atac)[, c("nearestGene", "distToTSS", "isPromoter")]))
```

```{r peaks-per-gene}
# How many ATAC peaks are annotated per gene?
peaks_per_gene <- table(rowData(sce_atac)$nearestGene)
summary(as.integer(peaks_per_gene))
```

---

# Quality control

## RNA quality control

Standard scRNA-seq QC: we flag cells with extreme total counts, extreme gene
detection, or high mitochondrial fraction.

```{r qc-rna}
# Get mitochondrial gene names directly from EnsDb
mt_gene_names <- genes_gr$gene_name[as.character(seqnames(genes_gr)) == "MT"]

# Then find which rows of sce_rna are mitochondrial
mt_genes <- which(rownames(sce_rna) %in% mt_gene_names)
cat("Mitochondrial genes found in sce_rna:", length(mt_genes), "\n")

rna_qc <- perCellQCMetrics(sce_rna, subsets = list(Mito = mt_genes))
rna_qc_filter <- quickPerCellQC(rna_qc, percent_subsets = "subsets_Mito_percent")

# Attach QC metrics for later visualisation
colData(sce_rna) <- cbind(colData(sce_rna), rna_qc)
                          
cat("RNA: removing", sum(rna_qc_filter$discard), "low-quality cells\n")
sce_rna <- sce_rna[, !rna_qc_filter$discard]
```

## ATAC quality control

For scATAC-seq we apply a minimum-fragments-per-cell filter.
A thorough pipeline would also compute TSS enrichment and nucleosome signal
(available in packages such as `ArchR` or `Signac`); here we keep it simple.

```{r qc-atac}
atac_qc <- perCellQCMetrics(sce_atac)

# Keep cells with at least 1 000 detected peaks
keep_atac <- atac_qc$detected >= 1000
cat("ATAC: removing", sum(!keep_atac), "low-quality cells\n")
colData(sce_atac) <- cbind(colData(sce_atac), atac_qc)
sce_atac <- sce_atac[, keep_atac]

```

## Intersect cells passing QC in both modalities

```{r intersect-cells}
common_cells <- intersect(colnames(sce_rna), colnames(sce_atac))
cat("Cells passing QC in both modalities:", length(common_cells), "\n")

sce_rna  <- sce_rna[,  common_cells]
sce_atac <- sce_atac[, common_cells]
```

---

# Normalisation

## RNA: scran pooling normalisation

```{r norm-rna, cache=TRUE}
# Quick clustering to guide pooling
clusters_rna <- quickCluster(sce_rna, min.size = 50)
sce_rna <- computeSumFactors(sce_rna, clusters = clusters_rna)
sce_rna <- logNormCounts(sce_rna)
```

## ATAC: TF-IDF normalisation

For scATAC-seq data the counts are near-binary (0/1 per peak per cell);
log-normalisation is inappropriate.
Term Frequency–Inverse Document Frequency (TF-IDF) is the standard approach:
it up-weights rare accessible peaks while down-weighting peaks open in
virtually every cell.

```{r norm-atac}
# TF-IDF implemented manually (or use ArchR / Signac for production)
tfidf <- function(mat) {
    # mat: features x cells sparse matrix
    # TF: normalise each cell to unit sum
    tf  <- t(t(mat) / colSums(mat))
    # IDF: log(1 + N / rowSums)
    idf <- log1p(ncol(mat) / (rowSums(mat > 0) + 1))
    tf * idf
}

counts_atac           <- counts(sce_atac)
tfidf_mat             <- tfidf(counts_atac)
assay(sce_atac, "tfidf") <- tfidf_mat
```

---

# Dimensionality reduction

## RNA: PCA and UMAP

```{r dimred-rna, cache=TRUE}
# Select highly variable genes
dec_rna <- modelGeneVar(sce_rna)
hvg_rna <- getTopHVGs(dec_rna, n = 2000)

sce_rna <- runPCA(sce_rna, subset_row = hvg_rna, ncomponents = 30)
sce_rna <- runUMAP(sce_rna, dimred = "PCA", n_neighbors = 30, min_dist = 0.3)
```

## ATAC: LSI (truncated SVD on TF-IDF) and UMAP

Latent Semantic Indexing (LSI) is the scATAC-seq analogue of PCA.
We use the TF-IDF matrix and `irlba`-based SVD.

```{r dimred-atac, cache=TRUE}
# Select top variable peaks by standard deviation of TF-IDF scores
peak_sds  <- rowSds(as.matrix(assay(sce_atac, "tfidf")))
top_peaks <- order(peak_sds, decreasing = TRUE)[seq_len(20000)]

# Truncated SVD (LSI)
lsi_res <- irlba::irlba(
    t(assay(sce_atac, "tfidf")[top_peaks, ]),
    nv = 30
)
lsi_embed <- lsi_res$u %*% diag(lsi_res$d)
rownames(lsi_embed) <- colnames(sce_atac)
colnames(lsi_embed) <- paste0("LSI", seq_len(ncol(lsi_embed)))

# Store in SCE and compute UMAP — drop first LSI component (correlated with
# sequencing depth, analogous to PC1 in bulk ATAC)
reducedDim(sce_atac, "LSI") <- lsi_embed[, -1]
sce_atac <- runUMAP(sce_atac, dimred = "LSI",
                    n_neighbors = 30, min_dist = 0.3,
                    name = "UMAP")
```

# Building the altSEE-ready SCE object

`altSEE` expects the ATAC experiment to live as an `altExp` inside the RNA
SCE.
The linking column, `nearestGene` in `rowData(altExp(...))`, can be used in 
altSEE to map peaks that belong to each gene selected in the RNA panel.

```{r build-sce}
# Subset ATAC peaks to those with a valid gene annotation 
valid_peak <- !is.na(rowData(sce_atac)$nearestGene)
sce_atac_annot <- sce_atac[valid_peak, ]
cat("Annotated peaks:", nrow(sce_atac_annot),
    "of", nrow(sce_atac), "total peaks\n")

# Nest ATAC as altExp inside RNA SCE 
altExp(sce_rna, "ATAC") <- sce_atac_annot

# Confirm structure
sce_rna
```

```{r confirm-altexp}
# The altExp is accessible via altExp()
altExp(sce_rna, "ATAC")
```

```{r check-linking}
# Linking field: nearestGene maps each ATAC peak to a gene name
head(rowData(altExp(sce_rna, "ATAC"))$nearestGene)
```



# Launching the altSEE app

## Configuring the initial panel layout

We configure a six-panel layout that illustrates the altSEE workflow:

1. **ReducedDimensionPlot** — UMAP of cells coloured by cell type from the main assay (RNA).
2. **AltReducedDimensionPlot** — UMAP of cells coloured by cell type from the AltExp (ATAC).
3. **RowDataTable** (RNA) — searchable gene table; clicking a row can drive
   the RNA panel and the ATAC panel plots (current default).
4. **ATAC-RowDataTable** (ATAC) — searchable gene table; clicking a row can drive the RNA panel and the ATAC panel plots.
3. **LinkedFeaturesAssayPlot** (RNA, main experiment) — expression of the
   selected gene across cells.
4. **LinkedFeaturesAssayPlot** (ATAC, altExp) — ATAC accessibility for all
   peaks annotated to a selected gene, displayed as a violin/dot plot per
   cell type.


```{r configure-app}
# Panel 1: UMAP coloured by cell type
rdp <- ReducedDimensionPlot(
    Type            = "UMAP",
    ColorBy         = "Column data",
    ColorByColumnData = "celltype",
    PanelWidth      = 6L
)
# Panel 2: AltExp UMAP coloured by cell type
ardp <- AltReducedDimensionPlot(
    Type            = "UMAP",
    ColorBy         = "Column data",
    ColorByColumnData = "celltype",
    PanelWidth      = 6L
)


# Panel 3: Searchable gene table
rdt <- RowDataTable(
    Selected   = "CD14",
    Search     = "CD14",
    PanelWidth = 6L
)

# Panel 4: Searchable AltExp table
ardt <- AltRowDataTable(
    PanelWidth = 6L
)

# Panel 5: altSEE RNA panel. LinkedFeaturesAssayPlot defines a Selection
# Experiment (altExp or main experiment) with a lookupColumn and a 
# Visualisation Experiment (altExp or main experiment) with a MapColumn. All
# features in the Visualisation experiment that map via the MapColum to the 
# selected feature in the Selection experiment via the lookupColumn are 
# visualised. Here we initialise the SelectionExperiment and the Experiment for
# visualisation both with the (Main). The plot then simply
# visualizes features of the Main experiment. In the app the user can then
# change the SelectionExperiment if they want to visualise the main exp through
# a linked altExp, e.g. ATAC altExp.
mlfap <- LinkedFeaturesAssayPlot(
    SelectionExperiment = "(Main)",
    LookupColumn        = "(rowname)",
    Experiment          = "(Main)",
    AltAssay           = "logcounts",
    MapColumn          = "(rowname)",
    XAxis              = "Column data",
    XAxisColumnData    = "celltype",
    ColorBy            = "Column data",
    ColorByColumnData  = "celltype",
    YAxisFeatureSource = "RowDataTable1", # driven by gene selection in table
    PanelWidth         = 6L
)


# Panel 6: altSEE ATAC panel — peaks linked to selected gene in main 
# experiment
alfap <- LinkedFeaturesAssayPlot(
    SelectionExperiment = "(Main)",
    LookupColumn        = "(rowname)",
    YAxisFeatureSource  = "RowDataTable1", # driven by gene selection in table
    Experiment          = "ATAC",
    AltAssay           = "tfidf",
    XAxis              = "Column data",
    XAxisColumnData    = "celltype",
    ColorBy            = "Column data",
    ColorByColumnData  = "celltype",
    MapColumn          = "nearestGene",
    PanelWidth         = 6L
)



initial_panels <- list(rdp, ardp, rdt, ardt, mlfap, alfap)
```

## Running the app

```{r launch-app, eval=FALSE}
app <- iSEE(
    se      = sce_rna,
    initial = initial_panels
)

if (interactive()) {
    shiny::runApp(app, launch.browser = TRUE)
}
```

The app opens with **CD14** pre-selected (a canonical monocyte marker).
The ATAC panel on the right immediately shows all peaks annotated to *CD14*,
coloured by cell type.


Try searching for **CD3D** (T-cell
marker), or **GNLY** (NK cell) in the gene table — the RNA and ATAC panels
update simultaneously.



# Session information

```{r session-info}
sessionInfo()
```

