---
title: altSee panels
author:
- name: Lieven Clement
  affiliation: 
  - Ghent University.
  email: lieven.clement@ugent.be
  
date: "`r BiocStyle::doc_date()`"
package: altSEE
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Panel universe}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{altSEE}
  %\VignetteKeywords{GeneExpression, proteomics, Visualization, QualityControl, GUI}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
library(BiocStyle)
```

```{r, eval=!exists("SCREENSHOT"), include=FALSE}
SCREENSHOT <- function(x, ...) knitr::include_graphics(x, dpi = NA)
```

# Overview

The `r Biocpkg("iSEE")` package provides a general and flexible framework for interactively exploring `SummarizedExperiment` objects.
However, in many cases, more specialized panels are required for effective visualization of specific data types.
The `r Biocpkg("altSEE")` package implements dedicated panel classes to visualise data from altExps in the `SingleCellExperiment` object that work directly in the `iSEE` application and can smoothly interact with other panels.
This allows users to quickly parametrize bespoke apps for their data to address scientific questions of interest that involves data in the mainExp and altExps.
We first load in the package:

```{r}
library(iSEEu)
library(altSEE)
library(msqrob2)
library(SingleCellExperiment)
library(scater)
```

# Differential abundance plots

To demonstrate the use of these panels we will perform a differental abundance
analysis on  of the CPTAC study 6 containing conditions A and B (Paulovich et al. 2010) from the example in the `cptac` vignette of the `msqrob2` package. 

```{r}
peptideFile <- "https://github.com/statOmics/msqrob2data/raw/refs/heads/main/dda/cptacAvsB_lab3/peptides.txt"
```

The quantitative columns in maxQuant start with "Intensity." (note the dot!).

```{r}
peptides <- read.delim(peptideFile)
names(peptides)
(quantCols <- grep("Intensity[.]", names(peptides), value = TRUE))
```

We now can import these data using the readQFeatures function. The peptide level expression data will be imported into R as an instance of class QFeatures named cptac with an assay named peptides. We also use the fnames argument to set the row-names of the peptides assay to the peptide sequences.

```{r}
qf <- readQFeatures(peptides, 
                    quantCols = quantCols, 
                    name = "peptides", 
                    fnames = "Sequence")
```

Below we update the sample (column) annotations to encode the two conditions, A and B, and the original sample numbers.

```{r}
qf$condition <- rep(c("A", "B"), each = 3)
qf$sampleId <- paste0(qf$condition, rep(7:9, 2))
colData(qf)
```
We will now perform the data processing and as in `msqrob2` vignette `cptac`

1. Replace missing values by `NA`.
2. Perform log2-transformation
3. Peptide filtering
4. Normalisation
5. Summarization to protein level

```{r warning=FALSE}
qf <- zeroIsNA(qf, "peptides") # 1. convert 0 to NA

qf <- logTransform(qf, base = 2, i = "peptides", name = "peptides_log") # 2.

qf <- filterFeatures(
qf, ~ Proteins != "" & ## 3. Remove failed protein inference
  !grepl(";", Proteins)) ##  3. Remove protein groups
qf <- filterFeatures(qf, ~ Reverse != "+" & ## 3. Remove decoys
                          Potential.contaminant != "+") ## 3. Remove contaminants
qf <- filterNA(qf, i = "peptides_log", pNA = .5) ## 3. Remove peptides that were only picked up in 1 or 2 samples

qf <- sweep( #Subtract log2 norm factor column-by-column (MARGIN = 2) 
  qf,
  MARGIN = 2,
  STATS = nfLogMedian(qf,"peptides_log"),
  i = "peptides_log",
  name = "peptides_norm"
) ## 4. Normalisation

qf <- aggregateFeatures(qf,
    i = "peptides_norm", 
    fcol = "Proteins", 
    na.rm = TRUE,
    name = "proteins"
) ## 5. summarisation to protein level
```

We model the data using msqrob2 and infer differential expression for the B-A group comparison

```{r warning=FALSE}
qf <- msqrob(
    qf,  
    i = "proteins",
    formula = ~ condition,
    robust = TRUE)

qf <- hypothesisTest(
  qf,
  i = "proteins",
  contrast = makeContrast("conditionB = 0",parameterNames = "conditionB")
  )
```

Convert `QFeatures` object to `SingleCellExperiment`

```{r}
sce <- getWithColData(qf, i ="proteins") |> 
   as("SingleCellExperiment")
rn <- rowData(sce) |> names()
rowData(sce) <- cbind(rowData(sce)[,!rn%in%c("msqrobModels","conditionB")], rowData(sce)$conditionB)

assayNames(sce) <- c("proteins","aggcounts")
```

Add reduced dim plot 

```{r}
sce <- sce |> 
  runMDS(exprs_values = 1) 
```

Add altExps with peptide dat

```{r}
altExp(sce,"peptides") <- qf[["peptides_norm"]]
assayNames(altExp(sce,"peptides")) <-"peptides_norm"

assay(altExp(sce,"peptides"),"peptides_log") <- assay(qf,"peptides_log")
```

## launch iSEE

```{r}
cptacAltExpApp <- iSEE(
  sce,
  initial = list(
    ReducedDimensionPlot(),
    VolcanoPlot(PanelId = 1L),
    RowDataTable(PanelId = 1L, 
                 RowSelectionSource = "VolcanoPlot1"),
    MainExpPlot(YAxisFeatureSource = "RowDataTable1", 
                 PlotType = "Scatter + lines", 
                 XAxis = "Column data",
                 XAxisColumnData = "sampleId"),
    AltExpPlot(YAxisFeatureSource = "RowDataTable1", 
                AltAssay = "peptides_norm",
                PlotType = "Scatter + lines", 
                XAxis = "Column data", 
                XAxisColumnData = "sampleId")
  )
)
```


To launch the app you can just enter its name `cptacAltExpApp` on the command line. 

```{r eval = FALSE}
cptacAltExpApp
```


```{r, echo=FALSE}
SCREENSHOT("screenshots/altExpApp.png")
```
